---
title: "New Zealand Agriculture and Population Trends"
---

New Zealand’s agricultural sector has long been central to its economy and identity.
In this analysis, I combine livestock counts with national population data to explore how animal numbers compare to human population growth over time.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(rvest)
library(scales)
```

Cleaning and Preparing the Agriculture Data

First, I load the TidyTuesday dataset and standardize minor naming inconsistencies.

```{r message=FALSE, warning=FALSE}
dataset <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-02-17/dataset.csv"
) |>
  mutate(
    measure = case_when(
      measure == "Total goats" ~ "Total Goats",
      measure == "Total Dairy Cattle (including Bobby Calves)" ~ "Total Dairy Cattle",
      measure == "Total Lambs Marked and/or Tailed" ~ "Total Lambs",
      TRUE ~ measure
    )
  )

animals <- dataset |> 
  filter(measure %in% c(
    "Total Sheep",
    "Total Beef Cattle",
    "Total Dairy Cattle",
    "Total Deer",
    "Total Goats", 
    "Total Lambs",
    "Total Pigs"
  ))
```

Adding New Zealand Population Data

Next, I scrape historical population data and align it by year.

```{r message=FALSE, warning=FALSE}
pop_nz <- read_html("https://commons.wikimedia.org/wiki/Data:New_Zealand_Total_Population.tab") |>
  html_elements("table") |> 
  html_table(header = TRUE) |> 
  as.data.frame() |> 
  slice(-(1:2)) |> 
  mutate(across(everything(), as.numeric)) |> 
  mutate(pop = y1 * 1000000)

animals_pop <- animals |> 
  left_join(pop_nz, by = c("year_ended_june" = "x"))
```

Livestock Counts vs Human Population

The following visualization compares major livestock categories with the total human population.
Population is scaled to align visually with animal counts and shown on a secondary axis.

```{r}
scale_factor <- 10

animals_pop |>
  ggplot(aes(x = year_ended_june, y = value, color = measure)) + 
  geom_line(linewidth = 1) + 
  scale_color_brewer(palette = "Set2") + 
  geom_line(aes(y = pop * scale_factor),
            linewidth = 1.3,
            color = "black") +
  scale_y_continuous(
    labels = comma,
    sec.axis = sec_axis(~ . / scale_factor,
                        name = "Human Population",
                        labels = comma)
  ) +
  labs(
    title = "New Zealand Livestock Numbers Compared to Human Population",
    subtitle = "Major livestock categories plotted alongside scaled population totals",
    x = "Year",
    y = "Livestock Count",
    color = "Livestock Category"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.margin = margin(5, 5, 5, 5),
    legend.position = "right"
  )
```